Loan Data Exploration by Ramiz Assaf
========================================================

> **Tip**: You will see quoted sections like this throughout the template to
help you construct your report. Make sure that you remove these notes before
you finish and submit your project!

> **Tip**: One of the requirements of this project is that your code follows
good formatting techniques, including limiting your lines to 80 characters or
less. If you're using RStudio, go into Preferences \> Code \> Display to set up
a margin line to help you keep track of this guideline!

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.
#install.packages("ggplot2", dependencies = T) 
#install.packages("knitr", dependencies = T)
#install.packages("dplyr", dependencies = T)
#install.packages("ggplot2", repos = 'http://cran.us.r-project.org')

library(ggplot2)
library(knitr)
library(dplyr)
```



```{r echo=FALSE, Load_the_Data}
# Load the Data
df_all <- read.csv('prosperLoanData.csv')
head(df_all)
```

This report explores a dataset containing a subset of 81 variables for \
approximately 113,937 loans.

# Univariate Plots Section


First we need to understand the type of each variable in the datafram using the str functions.
I have selected a number of variables. 

```{r echo=FALSE}
myvars <- c("Term", "CreditGrade", "BorrowerAPR", "EstimatedReturn", "EmploymentStatus", "IsBorrowerHomeowner", "IncomeRange",  "ProsperScore" , "LoanStatus", "BorrowerState")

df <-subset(df_all,  select = myvars)
str(df)
```

We have in our new subset all data types needed.
Now we are going to explore summaries for some variables to understand the \
data set better:

```{r echo=FALSE }
print(summary(df))
```

Term variable is interesting, most of the loan terms are 36 months. To \
look at it bettet I will apply table function
```{r echo=FALSE }
print(table(df$Term))
```

Majority of the loan terms is 36 months.

We can also take a look at the variable 'BorrowerAPR'

50% of the BorrowerAPR range between 15.6% and 28.3%. We have 25 missing  \
 values as well.

the variable "CreditGrade"" has a lot of blank rows. 

The variable EmployemntStatus has multiple Other values, I think it should be \
combined before processing.

With regard to EstimatedReturns Not all loans generated positive returns, some \
of them were losses. to count how many losses we can apply:

```{r echo=FALSE }
print(summary(df$EstimatedReturn > 0))
```

```{r echo=FALSE }
print(sort(table(df$BorrowerState)))
```

Most loans were given to Calafornia, Texas then New York and Florida.

We also can look the satus of employment for the borrower:

```{r echo=FALSE }
print(summary(df$EmploymentStatus))
```

most of the Borrowers are 'Employed', 2nd group is full-time, it is not clear to me if they indicate the same.

```{r echo=FALSE }
print(summary(df$LoanStatus))
```

Finally, we can look at the number of Recommendations the borrower provided \
for his loan.


Now we will plot some variables, we will try to observe patterns, strange \ 
behaviors, outliers...etc


```{r echo=FALSE, Univariate1_Plots}
df$CreditGrade2 <- factor(df$CreditGrade, levels=c("AA", "A", "B", "C", "D", "E", "HR", "NC", " "))
ggplot(aes(x = CreditGrade2), data = df)+ geom_bar() + xlab('Borrower Credit Grade') + labs(title='The count of different Credit Grades for Borrowers')
```

Majority withour a Credit Grade estimate.

Now we can plot other variables to get a good understanding of the data.

```{r echo=FALSE, Univariate_Plots}
ggplot(aes(x = CreditGrade2), data = df[(df$CreditGrade) != "", ])+ geom_bar(fill="blue") + xlab('Borrower Credit Grade') + labs(title='The count of different Credit Grades for Borrowers')
```

The highest Credit grades given are C, then D and B. 
I reorded them as level AA is higher than A


```{r echo=FALSE, Univariate2_Plots}
ggplot(aes(x =Term), data = df )+ geom_bar(fill='blue', color="darkblue", width = 8) + labs(title = ' Barchart of the loan TERM in months' ) + xlab('duration in months') + scale_x_discrete(limits = c(12,36,60))
```

Most of the terms are 36 month.

```{r echo=FALSE, Univariate3_Plots, message=FALSE}
ggplot(aes(x = BorrowerAPR, rm.na = TRUE ), data = df )+ geom_histogram(fill='orange', color = 'white', bins = 40) + labs(title = "A Histogram for the Annual Percentage Rate given for Prosper loans", x ='The Annual Percentage Rate (APR)') + scale_x_continuous(labels = scales::percent)
```


The majority of the APR values are between 10% and 35%. 




Now we can also look the revenue using the following code:


```{r echo=FALSE, Univariate6_Plots}
ggplot(aes(x= EstimatedReturn, rm.na = TRUE), data = df) + geom_histogram(fill ='darkblue', bins = 25) + labs(title =" The Loan Percentage Return", x ='Loan Percentage Return') + scale_x_continuous(labels = scales::percent, limits = c(-0.1, 0.25))
```

This histogram is far from being normal, it has a high peak. 


```{r echo=FALSE, Univariate7_Plots}
ggplot(aes(x= ProsperScore), data = df) + geom_bar(fill ='darkblue') + labs(title =" The Loan Percentage Return", x ='Prosper Score')
```


# Univariate Analysis

### What is the structure of your dataset?
The data has a dataframe sturcute with variable being in columns and recored \
reperesnt a loan case. 81 variables for approximately 113,937 loans.
We took a subset of 10 variables

Most borrowers are from California and Texas. Loans are usually 36 months.
The average Reurn is about 0.1 majority of loans are not classified with \
regard of risk. Of those classified, most classified categories are C and D.

### What is/are the main feature(s) of interest in your dataset?
The mean features is the Status of the Loan. Is is going to be paid? can we see \ 
patterns for when a loan is defaulted? we need to investigate

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

I choose the 10 variables after I examined the data a bit, I think I will \ 
be using them to predict the outcome of a loan to be paid on time or defaulted.

### Did you create any new variables from existing variables in the dataset?
On the contrary I removed some variables from the analysis as it would be hard \ 
explore them now. Plus at this point it is not needed, the data set has 81 \
variables covering a lot of ascpects. But indirectly I changed the order of \
factors

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?
I did it couple of times, where I had to remove NA instances
Some outliers exist, no need to remove them just now. they arw within limits. 

# Bivariate Plots Section



Would there be a relationship between the duration of the loan  \
and the BorowerAPR?

```{r echo=FALSE, Bivariate_Plots3}
ggplot(aes(y = BorrowerAPR, x = Term, rm.na = TRUE), data = df )+ geom_point(alpha = 1/100)+ labs(title="A Scatter plot between the Loan TERM and BorroweAPR", x = "Loan Duration TERM (in months)", y = "Borrower APR") + scale_y_continuous(labels=scales::percent)
```

I think at this point, there is not clear relationship \
between these two variables.

We need to replace the scatter with boxplot

```{r echo=FALSE, Bivariate2_Plots}
df[,'Term2']<-factor(df[,'Term'])
ggplot(aes(y = BorrowerAPR ,x = Term2,  fill = Term2), data = df )+ geom_violin(draw_quantiles = c(0.25,0.5,0.75))+ labs(title="A Scatter plot between the Loan TERM and BorroweAPR", x = "Loan Duration TERM (in months)", y = "Borrower APR") + scale_y_continuous(labels=scales::percent) + scale_x_discrete() + scale_fill_discrete(guide=FALSE)
```

The plot shows that loans of a term 12 months have an quasi-uniform \ 
distribution for the APR loans with a term 36 have the highest range, \
surprizingly the mode is about 36% while loans with term equal to \
60 the average is about 20%
\
```{r}
cor.test(df$Term, df$BorrowerAPR)
```

The correlation test indicates there is a very very weak inverse relationship.



Q. Would loans classified with higher risk, have higher APR?

```{r}
ggplot(aes(y = BorrowerAPR, x = ProsperScore), data = df )+ geom_point(alpha = 1/50)
```

The variable ProsperScore indicates a numerical estimation of the risk, \
the higher the number the lower the risk. the plot shows there is  \
a inverse relationship. to quantify this value we use the correlation test

```{r, echo=FALSE}
cor.test(df$BorrowerAPR,df$ProsperScore)
```

There is a strong relationship between the two variables, this makes perfect \ 
sense as the higher the risk usually higher the APR. Again the high risk \ 
is indicated with a low Prosper value.


Q. Would loans classified with higher risk for longer Terms, have higher APR ?

```{r, echo=FALSE}
ggplot(aes(y =BorrowerAPR , x = df$EmploymentStatus), data = df )+ geom_boxplot(aes(fill = df$Term2))+labs( title = "a box-plot Prosper Score vs Borrower APR for home owners and non- home owners" , y = " Borrower APR" ,x= "Prosper Score" , color='Is a Home Owner?' ) + theme(axis.text.x = element_text(angle = 90, hjust = 1))
```


The APR for "Not Employed" is higher, followed by Self-Employed. Furthermore, \
Terms with 12 months have gnerally higher APR.

we can at a different figure using the following code.
Instead of the Prosper Score we can look at the Credit Grade and if the\
borrower has a home or not.

The APR increase with lower classification.
Having a home generally increases the APR.


```{r, echo=FALSE}
ggplot(aes(y = BorrowerAPR, x = CreditGrade2), data = df )+ geom_boxplot(aes(color = df$IsBorrowerHomeowner), alpha = 1/10, fill = 'darkblue')+labs( title = "a Scatter plot Credit Grade vs Borrower APR for home owners and non- home owners" , y = " Borrower APR" ,x= "Credit Grade" , color='Is a Home Owner?')
```

It confirms the relationship as the dark blue dots are the home owners, \
most of them are classified AA to C, where as those who do not own a home \
are in most cases classified from D to NC.

Now can we say something about the BorrowerAPR vs Estimated return?

```{r echo=FALSE, bivar5_Plots, message=FALSE}
ggplot(aes(x = BorrowerAPR, y = EstimatedReturn, rm.na = TRUE ), data = df )+ geom_point(fill='orange', color = 'blue' ) + labs(title = "A Scatter plot for the Annual Percentage Rate vs Borrower APR ", y ='The Annual Percentage Rate (APR)' , x= 'Borrower APR')
```

It seems there is somehow some corrleation.But it also seems something else is \
affecting the relationship



Another thing comes into mind, what is the relationship between Risk and status \
of the loan?

```{r, echo=FALSE}
ggplot(aes(y = LoanStatus, x = CreditGrade2), data = df )+ geom_tile(fill ='blue' , alpha = 1/200)+labs( title = "a Scatter plot Credit Grade vs Loan Status" , y = " Loan Status" ,x= "Credit Grade" ) 
 
```

This figure shows some intresting patterns
If we look at the Defaulted row, we can notice the color is becoming darker, \
showing there is a strong relationship with the number defaulted with the \
Credit Grade.

The smae pattern is happenning to Chargedoff Loans i.e. the lower \
the Credit Grade the the more the chance of a Chargeoff

The Loans called Completed has an opposite relationship, the higher the Credit \
Grade the more loans are completed. The dots are fader with lower credit Grade.

Some parts are missing, there are not complete information about late Loans \
(or they are very few, we can know if we set alpha = 1)


The same plot can be done for the Employment variable. It goes as follows:

```{r, echo=FALSE}

ggplot(aes(y = LoanStatus, x = EmploymentStatus), data = df )+ geom_tile( alpha = 1/50)+labs( title = "a Scatter plot Employment Status vs Loan Status" , y = " Loan Status" ,x= "Employment Status" ) + theme(axis.text.x = element_text(angle = 90, hjust=0)) +
  scale_x_discrete(position = "bottom")
```


It seems Employment Status: Employed and Full-Time have the better \ 
chances of completing their loans. Employment status needs more \ 
examination as the variable has gaps that are not easy to explain \

To evalute if there is relationship betwee all studied variable. \
we use a matrix plot 


```{r echo= FALSE, message= FALSE}
#install.packages("GGally")
library('GGally')
drops <- c("BorrowerState", 'CreditGrade', 'EmploymentStatus', 'IsBorrowerHomeowner', 'IncomeRange', 'LoanStatus', 'BorrowerState', 'CreditGrade2', 'Term2', 'LoanStatusFactor')
df2 <- df[ , !(names(df) %in% drops)]
ggcorr(df2)

```


```{r echo= FALSE}
cor.test(df$BorrowerAPR,df$EstimatedReturn)

```

This is the highest correlation found.

# Bivariate Analysis


### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

Yes it did vary, for example, the complete loans are affected \ 
by the employement status.


### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

It is strange though that only employed category have most of the late loans. \
It doesn't make any senese. The data is not complete or misrecored.

### What was the strongest relationship you found?
Borrower APR and Estimated Return +0.79

# Multivariate Plots Section


```{r echo=FALSE, Multivariate4_Plots}
ggplot(aes(y = BorrowerAPR, x = CreditGrade2), data = df )+ geom_violin(aes(color = df$IsBorrowerHomeowner ), alpha = 1/100 )+labs( title = "a Scatter plot Credit Grade vs Borrower APR for home owners and non- home owners" , y = " Borrower APR" ,x= "Credit Grade" , color='Is a Home Owner?') 


```

It appears that owning a home doesn' affect greatly the BorrowerAPR
initillay I thought the home owners would enjoy a lower APR but now \ 
from the figure they actually dont

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

```{r echo=FALSE, Multivariate3_Plots}
ggplot(aes(y = BorrowerAPR, x = CreditGrade2), data = df )+ geom_point(aes(color = df$IsBorrowerHomeowner ), alpha = 1/100 )+ facet_wrap(df$IncomeRange)  + labs( title = "a Scatter plot Credit Grade vs Borrower APR for home owners and non- home owners" , y = " Borrower APR" ,x= "Credit Grade" , color='Is a Home Owner?') 


```

There is a reltionship, it can be seen that income between 75K and 99K are \ 
mostly homeowners, while income group over 100K are all homeowners.  \ 
Furthermore the higher the income the higher the classification, \ 
all of these variables have impact on whether a loan is paid on time \
or defaulted.

### Were there any interesting or surprising interactions between features?
Indeed, all unemployed got a Credit Grade of NA. Most of them without home \
Borrowes with Credit Grade NA exist at all income levels \


### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.
```{r echo=FALSE, MLR}

fit <- lm(EstimatedReturn ~ BorrowerAPR + IncomeRange + IsBorrowerHomeowner, data=df)
summary(fit) # show results


```


I will cancel out the "Income Range" variable

```{r echo=FALSE, MLR2}

fit <- lm(EstimatedReturn ~ BorrowerAPR + IsBorrowerHomeowner, data=df)
summary(fit) # show results


```

These two variable "BorroweAPR" and IsBorroweHomeownerTrue" are significant. 
It makes sense as the higher the BorrowerAPR it increases the revenues and with home owner being true it decreases the risk.

------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One
```{r echo=FALSE, Plot_One}
ggplot(aes(x = BorrowerAPR, rm.na = TRUE ), data = df )+ geom_histogram(fill='orange', color = 'blue' ) + labs(title = "A Hisogram for the Annual Percentage Rate given for Prosper loans", x ='The Annual Percentage Rate (APR)') 

```


This histogram shows the distribution of the BorrowersAPR, it is not exactly \
normally distributed, it has multi modes.

### Description One

In this plot it shows that the highest mode for the Annual Percentage rate APR \
is around 20%. the plot shows there are no outliers. This plot also shows that \
the APR is multimodeal distribution (more than two modes)





### Plot Two

```{r echo=FALSE, message= FALSE,  Plot_Two}
ggplot(aes(y = BorrowerAPR, x = CreditGrade2), data = df )+ geom_boxplot(aes(color = df$IsBorrowerHomeowner), alpha = 1/10, fill = 'darkblue')+labs( title = "a Boxplot plot Credit Grade vs Borrower APR for home owners and non- home owners" , y = " Borrower APR" ,x= "Credit Grade" , color='Is a Home Owner?')
```



### Description Two

This plot is very intersting, as one would think if the Borrower is not a \
homeowner they would be charged a higher APR, but what happens is the \
contrarty in most cases (except NC and NA).

It also should be noted that there are a lot of variation, many values are \ 
considered outliers .

### Plot Three

```{r echo=FALSE, Multivariate2_Plots}
ggplot(aes(y = BorrowerAPR, x = CreditGrade2 ), data = df )+ geom_boxplot(aes(color = df$IsBorrowerHomeowner ), alpha = 1/50 )+ facet_wrap(df$IncomeRange)  + labs( title = "a Boxplot Credit Grade vs Borrower APR for home owners and non- home owners for each income range" , y = " Borrower APR" ,x= "Credit Grade" , color='Is a Home Owner?') 

```

### Description Three

For each graph the lower the Credit Classificationm the higher the APR charged.
In all figures, if one has a home their APR is higher. (maybe it is to help \
those who dont have homes.)

------

# Reflection

This exploration is a lot of fun. it needs concentration and a lot of \
support from Udacity. I expected if you have a home you should have\
been with lesser risk. But results were not connected 100% to the issue.

One interesting question, how can we predict if the Borrower will default given\
the current information we have?

